New insights into the Weddell Sea ecosystem applying a quantitative network approach

### 2022-12-28 Manuscript/Manuscript.tex
Update bib

\title{New insights into the Weddell Sea ecosystem applying a network
approach}

@@

\affil[1]{Centro Austral de Investigaciones Científicas (CADIC-CONICET),
Ushuaia, Argentina}
\affil[2]{University of Helsinki, Helsinki, Finland}
\affil[*]{These authors contributed equally to this work.}

@@

\introduction[Introduction]

The objective of this work was to improve the knowledge on how the
Weddell Sea and the species therein may respond to perturbations from
ongoing environmental changes. To achieve this we: 1) estimated the
strength for each interaction in the Weddell Sea food web, 2)
characterised species considering weighted and unweighted properties,
and 3) analysed the species' role in the stability of the food web.

@@

The high Antarctic Weddell Sea shelf is situated between 74 and 78ºS
stretching approximately 450 km from East to West (Figure 1). Water
depth varies between 200 and 500 meters, and shallower areas are covered
by continental ice, which forms the coastline along the eastern and
southern part of the Weddell Sea. The shelf area contains a complex
benthic three-dimensional habitat with large benthic biomasses,
intermediate to high diversity in comparison to benthic boreal
communities and a spatially patchy distribution of organisms
\citep{Dayton1990, Teixido2002}.

@@

To estimate the strength of each pairwise interaction in the food web we
followed the methodology proposed by \citet{Pawar2012}. The minimum data
requirements are body mass of the consumer (predator) and resource
(prey), and the interaction dimensionality (ID) classified as 2 or 3
dimensions. The ID is defined as the dimension of the search space of
the predator, that is equivalent to the movement space of the prey.
Thus, the ID is classified as 2D when both predator and prey move in 2D
(e.g., both are benthic) or if a predator moves in 3D and a prey in 2D
(e.g., pelagic predator on benthic prey). The ID is classified as 3D
when both predator and prey move in 3D (e.g., both pelagic) or if the
predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic
prey) \citep{Pawar2012}. GATEWAy v.1.0 provides information on the mean
body mass for consumers and resources, except for `detritus' and
`sediment', and the dimensionality for the majority of the interactions,
though the latter is missing in some cases (924 interactions). To
complete the missing data on species `dimensionality', we used
information about the movement type of predators and prey included in
GATEWAy.

@@

We obtained estimates for resource density and the search rate from the
scaling relationships with the resource and the consumer mass,
respectively \citep{Pawar2012}. The coefficients of such relationships,
determined by ordinary least squares regression, vary with the
interaction dimensionality. On one hand, resource density scales with
resource mass as power-law with exponents in 2D and in 3D. Since mean
mass for resources `phytodetritus' and `sediment' were not available in
GATEWAy, we considered the body mass of the smallest phytoplankton
species (`Fragilariopsis cylindrus') as a proxy. This is justified by
the fact that `phytodetritus' and `sediment' are mainly composed of dead
or senescent phytoplankton reaching the seabed \citep{Wolanski2011}. On
the other hand, search rate scales with consumer mass as power-law with
exponents in 2D and in 3D.

@@

To characterise the role of each species in the food web, we considered
unweighted and weighted food web properties (Figure 2). Unweighted
properties are related to properties commonly used in qualitative food
web studies and only describe the presence or absence of interactions
without any information on strength between a pairwise species link
\citep{Martinez1991, Dunne2002, Borrelli2014}. In contrast, weighted
properties capture the importance of interaction strength.

To assess species roles as a function of the weighted food web, we
focused on mean interaction strength defined as the average strength of
all interactions for a given species. Further we calculated three
unweighted species properties: a) species degree, i.e., the sum of in-
and out-going interactions ; b) trophic level ; and c) trophic
similarity, i.e., the trophic overlap based on shared and unique
resources and consumers. These metrics were chosen to assess a species
role based on the unweighted food web. The species degree has often been
equated with species importance to the structure and functioning within
a food web, i.e.~perturbations to high-degree species may therefore have
more significant effects on the food web robustness to perturbations
than low-degree species
\citetext{\citealp{Dunne2002a}; \citealp[references
in][]{Cirtwill2018a}}. The trophic level offers information about how
important a species is to its biotic community, i.e., top predators and
primary producers are expected to have particularly large effects on the
rest of their communities through top-down and bottom-up control,
respectively \citep[references in][]{Cirtwill2018a}. Trophic similarity
is an index of trophic overlap considering the set of prey and predators
for a pair of species; it measures one of the most important aspects of
species' niches, the trophic niche, and functional aspects of
biodiversity \citep{Martinez1991, Williams2000}.

@@

After exploring the stability difference against the species properties
(Figure 5), we found that those species that generated a significant
impact on the stability of the food web were characterised by: 1) high
mean interaction strength; 2) mid to high trophic levels (TL
\textgreater{} 3.2); 3) relatively high number of interactions (Degree
\textgreater{} 25); and 4) mid to low trophic similarity (TS \textless{}
0.16). Habitat wise, species with a significant impact on the stability
were present in all habitats, except for the benthic realm. Table 2
shows these results for such species.

@@

Our analyses show that the distribution of species interaction strength
at the network level is asymmetric, i.e., the Weddell Sea food web
contains many weak interactions and only a few strong ones. This finding
is consistent with many previous theoretical and empirical studies
\citep[e.g.][]{McCann1998, Neutel2002, Emmerson2004, Wootton2005, Kortsch2021}.
The asymmetric distribution of interaction strength in food webs has
been interpreted as an explanation for the persistence of complex
communities in nature \citep{Bascompte2005, Allesina2015, Nilsson2016}.
Here we show that this pattern is also prevalent in one of the most
complex empirical (marine) food webs to date, comprising 490 species and
16041 predator-prey interactions. This finding reinforces the call for
the inclusion of interaction strength in food web studies to better
understand the ecosystem functioning, and species and whole network
responses to environmental perturbations.

@@

We employed a range of descriptors using both unweighted and weighted
food web properties to characterise the dynamic and multifaceted nature
of the Weddell Sea food web. Our results show a positive relationship
between interaction strength and trophic level, and between interaction
strength and species degree. Mean interaction strength increases with
trophic level and species degree. The former relationship might
contradict those studies that suggest that mid-trophic level species are
involved in the major pathways of energy flow in high-latitude marine
ecosystems
\citep{Pinkerton2014, Murphy2016, McCormack2020, Riccialdelli2020}. This
could be explained by the lack of species biomass data in our
interaction strength estimations; the methodology we applied here
\citep{Pawar2012} allows empirical data for the density of the resource
to be included, though this data is not for the majority of food web
species at the study site. On the other hand, the positive relationship
between interaction strength and degree reinforces the importance of
species with many interactions: species with high degree (hubs) have a
large impact on overall food web structure and functioning
\citep{Dunne2002a, Kortsch2015}. In the Weddell Sea, species with high
degree also tend to have high mean interaction strengths. This
information on the quantity and quality of interactions and its
relationship enables a robust assessment of the species' role in the
stability of the food web \citep{Cirtwill2018a}.

@@

Only a few species play a key role with respect to the Weddell Sea food
web stability, according to the mean maximum eingenvalue stability index
employed in this study. This is in concordance with other studies on
complex empirical food webs in marine ecosystems in the Arctic and other
locations in Antarctica \citep{Kortsch2015, Marina2018, Rodriguez2022}.
These key species are characterised by a particular set of food web
properties: high to mean interaction strength; mid to high trophic
level; relatively high number of interactions; and mid to low trophic
similarity. In a previous study on sequential extinction simulations for
the Weddell Sea food web, it was found that larger bodied-sized species
could be lost without causing a collapse of the network. A major caveat
of this finding, also recognised by the authors, was that population
dynamics were ignored and hence no top-down extinctions, or other
indirect effects, could occur. In our study we considered such top-down
effects by including information on the species interaction strength,
which is of paramount importance when analysing the response of
perturbations in ecological communities
\citep{McCann1998, Montoya2009, Novak2011}. Thus, our study suggests
that species with high mean interaction strength and high trophic level
need to be considered with particular attention when trying to predict
the effects of perturbations on the Weddell Sea ecosystem. This
conclusion is further reinforced by the finding that these species have
mid to low trophic similarity, which means that few other species of the
food web can occupy the same trophic role. In a review, it was
emphasised that polar pelagic communities are particularly sensitive to
changes due to a low functional redundancy at key trophic levels
\citep{Murphy2016}. Here we provide broader analyses of species impact
on food web robustness by including species from all habitats (benthic,
pelagic and land-based). This suggests that the sensitivity of marine
polar ecosystems to environmental perturbations is a concern also beyond
the pelagic realm.

@@

Our study goes beyond the current understanding of how species influence
ecosystem structure and stability in the Weddell Sea in particular and
in most polar regions in general \citep{Murphy2016, McCormack2021}. In
the same analysis we have integrated information about weighted
(interaction strength) and unweighted species properties, enabling a
more complete assessment of the species' role in the food web structure
and function, likely highlighting the ecological processes at play in
the Weddell Sea ecosystem.

>>>>

\title{New insights into the Weddell Sea ecosystem applying a
quantitative network approach}

@@

\affil[1]{Centro Austral de Investigaciones Científicas (CADIC-CONICET),
Ushuaia, Argentina}
\affil[2]{Spatial Foodweb Ecology Group, Department of Agricultural
Sciences, University of Helsinki, Finland}
\affil[*]{These authors contributed equally to this work.}

@@

\section{Introduction}

Food web analysis constitutes an important framework for understanding
ecological community structure and for conserving biodiversity through
ecosystem management \citep{Thompson2012}. Although topological food web
analysis, which considers only the presence and absence of predator-prey
interactions, provides important insights into the structure and
functioning of ecological communities
\citep[e.g.][]{Pascual2006, Kortsch2015, Marina2018, Cordone2020, Rodriguez2022},
more information on the nature of the trophic interactions is needed to
effectively charaterise ecosystem dynamics and stability
\citep[\citet{otherRefs}]{Kortsch2021}. This is a fundamental step for
providing assemessments on ecosystem vulnerability to environmental
pressures and for prioritising management actions. In this regard,
quantifying the strength of trohic interactions and species' roles
within the network are of paramount importance
\citep{Carrara2015, Allesina2015, Nilsson2016, Cirtwill2018a}.

We should maybe write a paragraph about weighted and unweighted
approaches here, where we aslo introduce the approach we used\ldots{}

\begin{quote}
{[}susanne{]} I moved the paragraph below. it makes more sense to first
talk about polar regions in genaral and then introduce the Weddell Sea.

Using a network approach, different types (e.g., terrestial, lake, etc.)
of food webs from various geographic locations are studied worldwide,
including marine polar food webs
\citep{Carscallen2012, Santana2013, Kortsch2019, Pecuchet2022}. Some of
these studies from the Arctic show how food web properties (e.g.,
connectance) are constrained by environmental factors such as sea ice
cover and seawater temperature \citep{Kortsch2019, Pecuchet2022}. In the
Southern Ocean, important insights have been gained into mechanisms of
energy flow, the relative importance and traits of individual species
and the influence of environmental variables (e.g., sea-ice) on the
structure of local food webs \citep{Cordone2020, Rossi2019}.
{[}susanne{]} what do you mean here: the relative importance and traits
of individual species? What important mechanisms? Can we be more
specefic? Can we also be more specific about how environmental variables
(e.g., sea-ice) influence the structure of local food webs in the SO?
\end{quote}

The Weddell Sea is expected to be one of the last regions of the
Southern Ocean to experience the consequences of climate change due to
its extensive ice cover and ocean currents \citep{Teschke2021} resulting
in less sea surface warming compared to other areas of the Southern
Ocean. Because of these environmental characteristics, the Weddell Sea
may serve as a refuge for Antarctic species which depend on sea ice
(e.g.~krill, emperor penguin, Weddell seal) or have low heat tolerance
(e.g., most notothenioid fishes) due to their adaptations to freezing
tempeartures \citep{Griffiths2017}. While essential large-scale
hydrodynamic relationships are relatively well-known for this region
\citep{deSteur2019}, information on the current distribution, abundance
and sensitivity to climate change is only partially known for a few
species (e.g., XXXXX) \citep{Houstin2022}. \textgreater{}
{[}susanne{]}it is not clear from the text why Weddell Sea is one of the
last regions of the SO to face the consequences of warming, be explicit
here, how are the currents affecting the Weddell Sea? \textgreater{}

Overall, the complexity the Weddell Sea food web is high and consists of
488 species and 16200 predator-prey interactions\citep{Jacob2011}. In an
attempt to better understand species role related to food web stability,
Jacob et al.~(2011) performed secondary extinction experiments and found
that the removal of small to medium-sized, and not large, organisms
caused a cascade of secondary extinctions. This findings highlighted the
relative importance of predators, rather than prey, in the architecture,
functioning and stability of the Weddell Sea food web, which coincides
with findings from recent meta-analyses in natural complex food webs
\citep{Brose2019, Perkins2022}. Other investigations have considered
this food web though in a meta-analysis context\ldots.{[}more text here,
WHAT DO THESE META\_ANALSYES STUDIES SHOW{]}
\citep{Brose2019, Perkins2022, Saravia2022}.

Several methodologies have been applied to estimate the interaction
strength in food webs, where the quantity and quality of data mostly
determine which approach is more convenient \citep{Berlow2004}.
Experimental methods combined with dynamic modelling
\citep{Emmerson2004, Carrara2015}, measurements of species abundances
trough time \citep{Fahimipour2014, Chang2021}, and/or estimation of
metabolic rates and biomass of all species in the community
\citep{Neutel2014}; all these methods require a large experimental
and/or sampling effort that restrict them to simple networks (circa 10
species). Other methods based on predator-prey body mass allometric
relationships \citep{Berlow2009, Pawar2012} could be applied to larger
networks due to less data requirements, although they might have less
power with increasingly complex networks unless more traits are included
in the estimations \citep{Jonsson2018}. For food webs composed by nearly
1000 species and about 20000 interactions only those methods with less
data requirements are feasible. On of these is the one proposed by
\citet{Pawar2012}, which allows to obtain interaction strength
estimations by combining data on body mass (of consumer -predator- and
resource -prey-) and interaction dimensionality (classified as 2 or 3
dimensions).

In this study, we aim to go beyond a purely topological
(presence/absence) assessment of who eats whom in the Weddell Sea
ecosystem by providing a quantitative analysis of the trophic
interaction network. We aim to analyse role for the structure and
stability of the food web. To achieve this, we: 1) estimated the
strength for each interaction in the Weddell Sea food web, 2)
characterised species role considering both weighted and unweighted
properties, and 3) analysed the species' role realted to the stability
of the food web. This is the first time that the interaction strength
was estimated for all the interactions mostly at the species level for
the Weddell sea food web

@@

The high Antarctic Weddell Sea shelf is situated between 74 and 78ºS,
stretching approximately 450 km in length from East to West (Figure 1).
Water depth varies between 200 and 500 meters, and shallower areas are
covered by continental ice, which forms the coastline along the eastern
and southern part of the Weddell Sea. The shelf area contains a complex
three-dimensional benthic habitat with large benthic biomasses,
intermediate to high diversity in comparison to benthic boreal
communities and a spatially patchy distribution of organisms
\citep{Dayton1990, Teixido2002}.

@@

To estimate the strength of each pairwise interaction in the food web,
we followed an approach proposed by \citet{Pawar2012}. The minimum data
requirements are body mass of the consumer (predator) and resource
(prey), and the interaction dimensionality (ID) classified as 2 or 3
dimensions. The ID is defined as the search space dimensions of the
predator, which is also equivalent to the movement space of the prey.
Thus, the ID is classified as 2D when both predator and prey move in 2D
(e.g., both are benthic) or if a predator moves in 3D and a prey in 2D
(e.g., pelagic predator on benthic prey). The ID is classified as 3D
when both predator and prey move in 3D (e.g., both pelagic) or if the
predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic
prey) \citep{Pawar2012}. GATEWAy v.1.0 provides information on the mean
body mass for consumers and resources, except for `detritus' and
`sediment', and the dimensionality for the majority of the interactions,
though the latter is missing in some cases (924 interactions). To
complete the missing data on species `dimensionality', we used
information about the movement type of predators and prey included in
GATEWAy.

@@

We obtained estimates for resource density and the search rate from the
scaling relationships with the resource and the consumer mass,
respectively \citep{Pawar2012}. The coefficients of such relationships,
determined by ordinary least squares regression, vary with the
interaction dimensionality. On one hand, resource density scales with
resource mass as power-law with different exponents in 2D and in 3D.
Since mean mass for resources `phytodetritus' and `sediment' were not
available in GATEWAy, we considered the body mass of the smallest
phytoplankton species (`Fragilariopsis cylindrus') as a proxy. This is
justified by the fact that `phytodetritus' and `sediment' are mainly
composed of dead or senescent phytoplankton reaching the seabed
\citep{Wolanski2011}. On the other hand, search rate scales with
consumer mass as power-law with exponents in 2D and in 3D.

@@

To characterise the role of each species in the food web, we considered
unweighted and weighted food web properties (Figure 2). Unweighted
properties are related to properties commonly used in qualitative food
web studies and only describe the presence or absence of interactions
without any information on strength between a pairwise species link
\citep{Martinez1991, Dunne2002, Borrelli2014}. In contrast, weighted
properties capture the importance of a trophic interaction by
considering its strength.

To assess species roles as a function of the weighted food web, we
focused on mean interaction strength defined as the average strength of
all interactions for a given species. Further we calculated three
unweighted species properties: a) species degree, i.e., the sum of in-
and out-going interactions ; b) trophic level ; and c) trophic
similarity, i.e., the trophic overlap based on shared and unique
resources and consumers. These metrics were chosen to assess a species
role based on the unweighted network. The species degree has often been
equated with species importance to the structure and functioning within
a food web, i.e.~perturbations to high-degree species may therefore have
more significant effects on the food web robustness to perturbations
than low-degree species
\citetext{\citealp{Dunne2002a}; \citealp[references
in][]{Cirtwill2018a}}. The trophic level offers information about how
important a species is to its biotic community, i.e., top predators and
primary producers are expected to have particularly large effects on the
rest of their communities through top-down and bottom-up control,
respectively \citep[references in][]{Cirtwill2018a}. Trophic similarity
is an index of trophic overlap considering the set of prey and predators
for a pair of species; it measures one of the most important aspects of
species' niches, the trophic niche, and functional aspects of
biodiversity \citep{Martinez1991, Williams2000}.

@@

After exploring the stability difference against the species properties
(Figure 5), we found that the species that generated a significant
impact on the stability of the food web were characterised by: 1) high
mean interaction strength; 2) mid to high trophic levels (TL
\textgreater{} 3.2); 3) relatively high number of interactions (Degree
\textgreater{} 25); and 4) mid to low trophic similarity (TS \textless{}
0.16). Habitat wise, species with a significant impact on the stability
were present in all habitats, except for the benthic realm. Table 2
shows the results for the species with highest impact on the food web
stability.

@@

Our analyses show that the distribution of species interaction strength
at the network level is asymmetric, i.e., the Weddell Sea food web
contains many weak interactions and only a few strong ones. This finding
is consistent with many previous theoretical and empirical studies
\citep[e.g.][]{McCann1998, Neutel2002, Emmerson2004, Wootton2005, Kortsch2021}.
The asymmetric distribution of interaction strength in food webs has
been interpreted as an explanation for the persistence of complex
communities in nature \citep{Bascompte2005, Allesina2015, Nilsson2016}.
Here we show that this pattern is also prevalent in the Weddell Sea, one
of the most complex food webs to date, comprising 490 species and 16041
predator-prey interactions. This finding is in someway validating the
method we used, validation of allometric methods of IS estimation not
including interaction dimensionality has been performed for microcosmos
of relatively few species \citep{Jonsson2018}.

@@

We employed a range of descriptors using both unweighted and weighted
food web properties to characterise the dynamic and multifaceted nature
of the Weddell Sea food web. Our results show a positive relationship
between interaction strength and trophic level, and between interaction
strength and species degree. Mean interaction strength increases with
trophic level and species degree. The former relationship contradicts
studies that suggest that mid-trophic level species (which, krill???
examples are always good) are involved in the major pathways of energy
flow in high-latitude marine ecosystems
\citep{Pinkerton2014, Murphy2016, McCormack2020, Riccialdelli2020}.
However, this finding could also be explained by the lack of species
biomass information in the calculation of interaction strength using the
methodology we applied \citep{Parwar2012}. The methodlogy by Parwar et
al.(2012) allows information on species density/biomass to be included,
but this type of data was not available for the majority of species in
our food web. In the Weddell Sea, species with high degree also tend to
have high mean interaction strengthsThe positive relationship between
interaction strength and species degree reinforces the central role of
species with many interactions: species with a high degree (hubs) have a
large impact on overall food web structure and functioning
\citep{Dunne2002a, Kortsch2015}. This information on the quantity and
quality of interactions and its relationship enables a robust assessment
of the species' role in the stability of the food web
\citep{Cirtwill2018a}.

@@

Only a few species play a key role with respect to the Weddell Sea food
web stability, according to the mean maximum eigenvalue stability index
employed in this study. This is in concordance with other studies on
complex empirical food webs in marine ecosystems in the Arctic and other
locations in Antarctica \citep{Kortsch2015, Marina2018, Rodriguez2022}.
These key species are characterised by a particular set of food web
properties: high to mean IS; mid to high trophic level; a relatively
high number of interactions; and mid to low trophic similarity. In a
previous study on sequential extinction simulations for the Weddell Sea
food web \citep[!!]{CITA}, it was found that larger bodied-sized species
could be lost without causing a collapse of the network. A major caveat
of this finding, also recognised by the authors, was that population
dynamics were ignored and hence no top-down extinctions, or other
indirect effects, could occur. In our study we considered such top-down
effects by including information on the species IS, which is of
paramount importance when analysing the response of perturbations in
ecological communities \citep{McCann1998, Montoya2009, Novak2011}. Thus,
our study suggests that species with high mean IS and high trophic
levels need to be considered with particular attention when trying to
predict the effects of perturbations on the Weddell Sea ecosystem. This
conclusion is further reinforced by the finding that these species have
mid to low trophic similarity, which means that few other species of the
food web can occupy the same trophic role. In a review, it was
emphasised that polar pelagic communities are particularly sensitive to
changes due to a low functional redundancy at key trophic levels
\citep{Murphy2016}. Here we provide broader analyses of species impact
on food web robustness by including species from all habitats (benthic,
pelagic and land-based). This suggests that the sensitivity of marine
polar ecosystems to environmental perturbations is a concern also beyond
the pelagic realm.

@@

Our study goes beyond the current understanding of how species influence
ecosystem structure and stability in the Weddell Sea in particular and
in most polar regions in general \citep{Murphy2016, McCormack2021}. In
the same analysis we have integrated information about weighted
(interaction strength) and unweighted species properties, enabling a
more complete assessment of the species' role in the food web structure
and function.
